### 2014-09-10 paper.tex
revised based on comments from Hans and Jerry

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Splice variants play an important role in biological systems especially in the
immune system and the central nervous system. An expression profile of splice
variants has been shown to be a better signature of some diseases than an
overall gene expression profile~\cite{zhang2013isoform}. However, for most
organisms, available gene models such as Ensembl may not include all splice
variants expressed in RNA-Seq data. Reference-guided (e.g.
Cufflinks~\cite{Trapnell:2010kd}) and de novo assembly (e.g.
Velvet/Oases~\cite{Schulz:2012je}) methods are available for building gene
models from RNA-Seq reads. However, we found that both types of method recovered
distinct set of splice variants.  In this article, we introduce a pipeline that
can combined transcripts from {\em de novo} assembly and other gene models to
constructs gene models that include more splice variants. We also described a
method called local assembly that we used to enhance the assembly of splice
variants to recover splice variants not found by Cufflinks and {\em de novo}
assembly.  Combination of Cufflinks gene models and transcripts from assembly
include more splice variants, increasing sensitivity of alternative splicing
study.
% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.

@@

Despite the small size of sequencing reads, several studies have
detected novel splice junctions based on alignment of reads spanning
across putative exon junctions.  To map reads across exon junctions,
reads are split into two parts and each part is mapped to the genome
independently (@cite).  A splice junction is then identified based on
alignments of each half of a read that fall between two exons at
exon-intron boundaries.  With this approach, Wang \emph{et al} have
identified a large number of splice junctions that are not annotated
from human cell lines (HUVEC and NHEK)\cite{Wang:2011jq}.  These novel
splice sites include both canonical and non-canonical splice sites.
Approximately, $46\%-75\%$ of canonical splice sites are supported by
ESTs.  Novel splice junctions have different levels of read coverage
suggesting that both high- and low-expressed isoforms are unannotated.

Using a similar approach, Pickrell {\em et al} identified more than
150,000 novel canonical splice junctions in lymphoblastoid cells.  The
study also shows that the number of unannotated splice junctions
varies among cells from different human tissues, which suggests
tissue-specific expression of isoforms\cite{Pickrell:2010gt}. 
% A majority of unannotated isoforms found in both studies are from alternative
% splicing of annotated exons. Only a small fraction of isoforms contain
% previously unknown exons.
% @CTB is this last point important?

Several tools have been developed not only to detect novel splice
junctions but also to reconstruct full-length isoforms from short
reads without using prior gene annotations.  These tools are especially
useful for transcriptome analysis of organims with incomplete gene
annotations.  Cufflinks\cite{Trapnell:2010kd} relies on splice
junctions detected from Tophat\cite{Trapnell:2009dp}, a read aligner
that can align reads across putative exon junctions, to reconstruct a
full-length transcript.  Cufflinks identified 12,712 novel isoforms,
of which 7,395 ($58\%$) contain novel splice junctions in mouse
myoblast cell lines.  Guttman {\em et al} used Scripture, a tool
employing similar mapping-based approach, to reconstruct a full-length
transcripts from mouse RNA-Seq data and discovered approximately 490
novel alternative isoforms in lincRNA loci, which are expressed in any
of the three different cell types\cite{Guttman:2010io}.

@@

A number of {\em de novo} assemblers have been used to reconstruct
transcripts from RNA-Seq data in many studies.
Trinity\cite{Grabherr:2011jb}, was successfully used to reconstruct
transcripts from yeast and mouse datasets.  It was also shown that
Trinity detects a unique set of novel splice junctions not detected by
Cufflinks or Scripture.  This suggests that a {\em de novo} assembly
approach is capabale of increasing sensitivity of detecting
alternative isoforms of a mapping-based method.
Trans-Abyss\cite{Robertson:2010ih} and Oases\cite{Schulz:2012je} are
extensions of the Abyss\cite{Simpson:2009iv} and
Velvet\cite{Zerbino:2008vu,Zerbino:2009jp} genome assemblers that are
tuned to work with RNA-Seq data.  These assemblers are comparable at
reconstructing existing and novel alternative isoforms with a slightly
different sensitivity and specificity.  However, Oases with Oases-M
has been shown to be superior than other {\em de novo} assemblers at
discovering isoforms in human and mouse\cite{Schulz:2012je}.

In this study, we present a pipeline that uses {\em de novo} assembly
to reconstruct alternative isoforms in RNA-Seq data from chickens.  We
apply a technique we call ``local assembly'' that enhances the sensitivity
of alternative isoform detection by Oases.  The results show that the
pipeline can detect more isoforms than Oases-M and can detect isoforms
not found by Cufflinks.  We also showed that transcripts reconstructed
from {\em de novo} assembly and mapping-based approach can be merged
to build more complete gene models.

@@

More importantly, we also mapped paired-end reads from technical replicates to
the same gene models, and found that more than 74\% of the paired-ends mapped
concordantly the gene models.  Most of the reads that did not map were either
highly erroneous or contained low-complexity artifactual sequence that probably
originated from sample processing and reverse transcription (@cite).

@@

Thus the merged gene models produced by Gimme represent the significant majority
of the assemblable data.

@@

Of 105,461 splice junctions from the gene models, 2,057 (2\%) junctions have no
spliced reads and only 4,626 (4.4\%) junctions have fewer than 4 spliced reads
(Fig.~\ref{cdf_splice}).  Note that 710 junctions are in random chromosomes,
which may have a great number of genome missassemblies. The number of junctions
outside random chromosomes with no reads is 1,347/99,986 (1.3\%).

% @CTB how do you get junctions with no spliced reads!?
% @LP our models have many 2,400 more junctions in random chromosomes than
% cufflinks and 710 of no reads junctions are from random chromosomes. It is
% likely that those junctions are from missassemblies.
% @LP Another reason is it has something to do with variation in read
% mapping between Tophat and bowtie.  Tophat maps spliced reads with less
% stringency because it has to allow gap alignment and it uses -v mode to map
% spliced reads. On the other hand, the default mode for bowtie is -n mode,
% which is the one I used.
% @LP I used a parameter that set max alignments to 100, only one alignment will
% be reported if there are more than 100 alignments. Maybe I should let
% bowtie report all alignments instead.

More than 95\% of our predicted splice junctions have a coverage
of 4 or higher in our combined mRNAseq data sets, suggesting that they are
real splice junctions.

% @CTB I'm slightly bugged by this statement.  Are we right?!
% @LP the reason I use 4 as a cutoff because there are four samples and if at
% least one read from each sample map to the junction, it could be real.
% Apparently, it is not always the case, but that is my simple logic.

@@

Of the 105,461 splice junctions in our gene models, 83,560 (79.2\%) are
supported by ESTs or mRNAs from Genbank~\ref{chick_est_venn}.  This is
especially surprising since our mRNAseq data is from spleen, and most of the
publicly available ESTs or mRNAs are from other tissues. 16,909 splice junctions
are found by either Cufflinks or our gene models suggesting a variation in
sensitivity between two methods. Note that this cross-validation suggests that
the 12,188 novel splice junctions {\em not} seen in publicly available ESTs and
mRNAs are also likely to be real splice junctions from spleen.

@@

As shown above, Cufflinks and the Gimme method detect a number of distinct but
equally valid splice junctions, which suggests that we could obtain greater
sensitivity to exon-exon junctions in our gene models by merging both sets of
predictions.  We therefore used Gimme to incorporate the Cufflinks models to
global and local assembly Table.~\ref{genes_transcripts}.  This resulted in a
decreased number of total genes, suggesting that some fragmented genes were
merged together to form more complete gene structures (e.g. see
Fig.~\ref{}). The merged gene models recover 49.3\% and 56.2\% of
splice junctions from ESTs and ENSEMBL respectively
(Figure~\ref{combined_venn}), which is about 10\% greater than that from
corresponding unmerged gene models.

@@

For organisms with a reference genome and Ensembl annotation, building gene
models from RNA-Seq reads using Cufflinks with Ensembl gene models as a
reference guide seem to be a preferable method for gene and isoform expression
analysis.  However, we have shown that Ensembl models in chicken are missing a
substantial number of splice variants found in ESTs and mRNAs and Cufflinks does
not recover all splice variants.  Previous study by Schulz \emph{et
al.}\cite{Schulz:2012je} demonstrates that Cufflinks outperformed \emph{de novo}
assembly in detecting genes and isoforms in Ensembl annotations in mouse.
However, we found that Cufflinks almost exclusively finds splice junctions in
mouse Ensembl gene models when our gene models include splice junctions not in
Ensembl gene models but supported by ESTs (Figure~\ref{mus_venn}).

To build mouse Cufflinks gene models, we used Tophat default parameters claimed
to be fine-tuned for mammalian RNA-Seq reads, but Cufflinks still poorly
detected splice junctions not in Ensembl.  This suggests that the parameters may
have been set based on mammalian Ensembl models, which could limit the
efficiency of finding novel splice variants in other non-mammal organisms.  In
contrast, {\em de novo} assembly does not rely on species-specific parameters.
It is, therefore, recommended that a combination of Cufflinks models and
transcripts from de novo assembly should be used to build gene models that
include more splice variants.

We have also shown that the local assembly could greatly increase sensitivity
of splice variant detection. However, how the local assembly affect the assembly
is not clearly understood.  We speculated that read with multiple alignments may
play a significant role in enhancing the assembly of splice variants.
Understanding of the mechanisms may lead to a technique that can be applied to
find splice variants in organisms that lack a reference genome.

Both Cufflinks and our pipeline relies heavily on a reference genome, therefore,
the quality of the genome will greatly affect the quality of the gene models.
In this study gene models were built from chicken genome version 2.1 (galGal3),
which contains a considerable number of sequence duplications and missassemblies
that supposed to be eliminated in the latest version of genome assembly
(galGal4).  Duplications and misassemblies lead to false splice junctions, which
in turn produce a large number of splice variants as observed in some
chromosomes.

@@

\subsection*{Mapping reads to The Reference Genome and Gene Models}

Single and paired-end reads were mapped to chicken genome by
Tophat2\cite{Trapnell:2009dp} release 2.0.0 using default parameters without
annotations.  All reads were mapped to cDNA sequences derived from gene models
by Bowtie2\cite{Langmead:2009fv} with default parameters \texttt{(n=2, l=28,
e=70, k=1)}.  Reads from mouse dataset were mapped to mouse genome (mm9)
downloaded from Tophat website \texttt{http://tophat.cbcb.umd.edu}.

@@

Reads from each dataset were first assembled separately in global assembly
without using a reference genome.  In contrast, reads from each dataset were
first mapped to the chicken genome using Tophat2.  Then only reads mapped to the
genome were assembled by chromosomes in local assembly
(Fig.~\ref{local_assembly}).  Global and local assembly was performed using
Velvet version 1.2.03\cite{Zerbino:2008vu} with default parameters except for
hash length (k-mer).  A range of k-mer length from 21-31 was used to assemble
reads from chicken data and k-mer length 27 was used to assemble reads from
mouse data.  Lastly, transcripts from both methods were assembled by Oases
version 0.2.06\cite{Schulz:2012je}.
A poly-A tail, short transcripts and transcripts with low complexity are removed
by seqclean~\cite{seqclean} with default parameters.  Redundant transcripts are
removed by cd-hit-est from CD-HIT suite\cite{Li:2006hr}.  A substantial number
of transcripts are removed at this step, which facilitates gene models
construction process.

@@

Figure~\ref{overall_pipeline} depicts an overall gene model construction
pipeline.  Transcripts of all datasets from local and global assembly were
mapped to the chicken genome using BLAT\cite{Kent:2002tv} \texttt{(-t=dna -q=dna
-noHead -out=psl -mask=lower -extendThroughN -dots=1000)}.  Alignments and gaps
from BLAT outputs are considered exons and introns respectively.  Optionally,
data from other sources (ESTs, RefGenes, etc.) can be incorporated with
transcripts from assembly to improve gene models.  All transcripts are then
assembled using Gimme, a program that assembles transcripts based on their
alignments to the reference genome.  An algorithm for assembling transcripts is
described below.  A maximum set of transcripts obtained from Gimme are then
reduced to only a minimum set of transcripts that contain all splice junctions
and untranslated regions (UTRs).  After that, transcripts that are highly
similar ($>99\%$) are clustered and removed by CD-HIT version
4.5.6\cite{Li:2006hr}.  Parameters for clustering are \texttt{-n 10, -r 0 and -c
0.99}.  Only a representative of each cluster is kept in gene models.

@@

A gene model can be represented as a splice graph composed of exons as nodes and
introns as edges.  However, transcripts of the same gene vary in size and
structure depending on the expression level and a hash length number used in
assembly.  Furthermore, incomplete exons and fragmented transcripts complicate
the construction of a splice graph.  In this study, we developed an algorithm
that handles incomplete exons and fragmented transcripts and constructs a
maximum assembly of gene models.

The algorithm first builds an intron graph using introns as nodes.  Each intron
contains exons whose one of their splice sites perfectly match intron
boundaries.  Exons are considered incomplete and eliminated if they locate at
the $3'$ or $5'$ end of the transcripts and they are not the largest exons (exon
3a and 3b in Fig.~\ref{algorithm}).  Transcripts were then grouped into the same
gene if they have at least one intron or exon in common.  Then, a splice graph
composed of exons is created and structures of isoforms are derived from
traversing paths in the splice graph.  Gimme is open-source and available at
\texttt{https://github.com/ged-lab/gimme}.

@@

We employed ESTScan version 3.0.3 to translate protein sequences from our gene
models.  The matrix used for building Hidden Markov model was built from chicken
reference cDNA sequences using tools from ESTScan. Only protein sequences longer
than 50 bp are included in the analysis.

@@

\subsection*{Sequence assembly using Cufflinks}
Reads are mapped to a genome sequence using Tophat.  Gene models are built from
each datasets by Cufflinks 2.0.0\cite{Trapnell:2010kd}.  All gene models are
then merged together using Cuffmerge.

\subsection*{Expressed sequence tags and Genbank mRNA}
Expressed sequence tags (ESTs) and mRNAs were downloaded from UCSC genome
website.  The database was updated from GENBANK on 1 January 2014.  Sequences
were aligned to a chicken genome using BLAT.

@@

\section*{Figure Legends}
\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{overall_pipeline.pdf}
\end{center}
\caption{
{\bf Gene model construction pipeline.} Transcripts are obtained from two
assembly methods -- global and local assembly.  Transcripts are aligned to a
chicken genome by BLAT\@. Gimme then constructs gene models based on alignments
of transcripts.
}
\label{overall_pipeline}
\end{figure}

>>>>

% Please keep the abstract between 250 and 300 words
\section*{Abstract}
Splice variants play an important role in biological systems, especially in the
immune system and the central nervous system. An expression profile of splice
variants has been shown to be a better signature of some diseases than an
overall gene expression profile~\cite{zhang2013isoform}. However, for most
organisms, available gene models such as Ensembl may not include all splice
variants expressed in RNA-Seq data. Reference-guided (e.g.
Cufflinks~\cite{Trapnell:2010kd}) and de novo assembly (e.g.
Velvet/Oases~\cite{Schulz:2012je}) methods are available for building gene
models from RNA-Seq reads. However, we found that both types of method recovered
distinct sets of splice variants.  In this article, we introduce a pipeline that
can combine transcripts from {\em de novo} assembly and other gene models to
constructs gene models that include more splice variants. We also described a
method called local assembly that we used to enhance the assembly of splice
variants to recover splice variants not found by Cufflinks and {\em de novo}
assembly.  Combination of Cufflinks gene models and transcripts from assembly
include more splice variants, increasing sensitivity of alternative splicing
study.
% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.

@@

Despite the small size of sequencing reads, several studies have
detected novel splice junctions based on alignment of reads spanning
putative exon junctions.  To map reads across exon junctions,
reads are split into two parts and each part is mapped to the genome
independently (@cite).  A splice junction is then identified based on
alignments of each half of a read that falls between two exons at
exon-intron boundaries.  With this approach, Wang \emph{et al} have
identified a large number of splice junctions that are not annotated
from human cell lines (HUVEC and NHEK)\cite{Wang:2011jq}.  These novel
splice sites include both canonical and non-canonical splice sites.
Approximately, $46\%-75\%$ of canonical splice sites are supported by
ESTs.  Novel splice junctions have different levels of read coverage
suggesting that both high- and low-expressed isoforms are unannotated.

Using a similar approach, Pickrell {\em et al}~\cite{Pickrell:2010gt}
identified more than 150,000 novel canonical splice junctions in lymphoblastoid
cells.  The study also shows that the number of unannotated splice junctions
varies among cells from different human tissues, which suggests tissue-specific
expression of isoforms\cite{Pickrell:2010gt}.
% A majority of unannotated isoforms found in both studies are from alternative
% splicing of annotated exons. Only a small fraction of isoforms contain
% previously unknown exons.
% @CTB is this last point important?

Several tools have been developed not only to detect novel splice
junctions but also to reconstruct full-length isoforms from short
reads without using prior gene annotations.  These tools are especially
useful for transcriptome analysis of organims with incomplete gene
annotations.  Cufflinks\cite{Trapnell:2010kd} relies on splice
junctions detected from Tophat\cite{Trapnell:2009dp}, a read aligner
that can align reads across putative exon junctions, to reconstruct a
full-length transcript.  Cufflinks~\cite{} identified 12,712 novel isoforms,
of which 7,395 ($58\%$) contain novel splice junctions in mouse
myoblast cell lines.  Guttman {\em et al} used Scripture, a tool
employing a similar mapping-based approach, to reconstruct a full-length
transcripts from mouse RNA-Seq data and discovered approximately 490
novel alternative isoforms in lincRNA loci, which are expressed in any
of the three different cell types\cite{Guttman:2010io}.

@@

A number of {\em de novo} assemblers have been used to reconstruct
transcripts from RNA-Seq data in many studies.
Trinity\cite{Grabherr:2011jb} was successfully used to reconstruct
transcripts from yeast and mouse datasets.  It was also shown that
Trinity detects a unique set of novel splice junctions not detected by
Cufflinks or Scripture.  This suggests that a {\em de novo} assembly
approach is capable of increasing sensitivity of detecting
alternative isoforms over a mapping-based method.
Trans-Abyss\cite{Robertson:2010ih} and Oases\cite{Schulz:2012je} are
extensions of the Abyss\cite{Simpson:2009iv} and
Velvet\cite{Zerbino:2008vu,Zerbino:2009jp} genome assemblers that are
tuned to work with RNA-Seq data.  These assemblers are comparable at
reconstructing existing and novel alternative isoforms with a slightly
different sensitivity and specificity.  However, Oases with Oases-M
has been shown to be superior to other {\em de novo} assemblers at
discovering isoforms in human and mouse\cite{Schulz:2012je}.

In this study, we present a pipeline that uses {\em de novo} assembly
to reconstruct alternative isoforms in RNA-Seq data from chickens.  We
apply a technique we call ``local assembly'' that enhances the sensitivity
of alternative isoform detection by Oases.  The results show that the
pipeline can detect more isoforms than Oases-M and can detect isoforms
not found by Cufflinks.  We also showed that transcripts reconstructed
from {\em de novo} assembly and mapping-based approaches can be merged
to build more complete gene models.

@@

More importantly, we also mapped paired-end reads from technical replicates to
the same gene models, and found that more than 74\% of the paired-ends mapped
concordantly the gene models.  Most of the reads that did not map were either
highly erroneous or contained low-complexity artifactual sequence that probably
originated from sample processing and reverse transcription (@cite).  Thus the
merged gene models produced by Gimme represent the significant majority of the
assemblable data.

@@

@@

Of 105,461 splice junctions from the gene models, 2,057 (2\%) junctions have no
spliced reads and only 4,626 (4.4\%) junctions have fewer than 4 spliced reads
(Fig.~\ref{cdf_splice}).  Note that 710 junctions are in chrUn\_random contigs,
which may have a great number of genome missassemblies. The number of junctions
outside random chromosomes with no reads is 1,347/99,986 (1.3\%).
% @CTB how do you get junctions with no spliced reads!?
% @LP our models have many 2,400 more junctions in random chromosomes than
% cufflinks and 710 of no reads junctions are from random chromosomes. It is
% likely that those junctions are from missassemblies.
% @LP Another reason is it has something to do with variation in read
% mapping between Tophat and bowtie.  Tophat maps spliced reads with less
% stringency because it has to allow gap alignment and it uses -v mode to map
% spliced reads. On the other hand, the default mode for bowtie is -n mode,
% which is the one I used.
% @LP I used a parameter that set max alignments to 100, only one alignment will
% be reported if there are more than 100 alignments. Maybe I should let
% bowtie report all alignments instead.
More than 95\% of our predicted splice junctions have a coverage
of 4 or higher in our combined mRNAseq data sets, suggesting that they are
real splice junctions.
% @CTB I'm slightly bugged by this statement.  Are we right?!
% @LP the reason I use 4 as a cutoff because there are four samples and if at
% least one read from each sample map to the junction, it could be real.
% Apparently, it is not always the case, but that is my simple logic.

@@

Of the 105,461 splice junctions in our gene models, 83,560 (79.2\%) are
supported by ESTs or mRNAs from Genbank (Fig.~\ref{chick_est_venn}).  This is
especially surprising since our mRNAseq data is from spleen, and most of the
publicly available ESTs or mRNAs are from other tissues. 16,909 splice junctions
are found by either Cufflinks or our gene models suggesting a variation in
sensitivity between two methods. Note that this cross-validation suggests that
the 12,188 novel splice junctions {\em not} seen in publicly available ESTs and
mRNAs are also likely to be real splice junctions from spleen.

@@

As shown above, Cufflinks and the Gimme method detect a number of distinct but
equally valid splice junctions, which suggests that we could obtain greater
sensitivity to exon-exon junctions in our gene models by merging both sets of
predictions.  We therefore used Gimme to incorporate the Cufflinks models to
global and local assembly, Table.~\ref{genes_transcripts}.  This resulted in a
decreased number of total genes, suggesting that some fragmented genes were
merged together to form more complete gene structures (e.g. see
Fig.~\ref{}). The merged gene models recover 49.3\% and 56.2\% of
splice junctions from ESTs and ENSEMBL respectively
(Figure~\ref{combined_venn}), which is about 10\% greater than that from
corresponding unmerged gene models.

@@

For organisms with a reference genome and Ensembl annotation, building gene
models from RNA-Seq reads using Cufflinks with Ensembl gene models as a
reference guide seems to be a preferable method for gene and isoform expression
analysis.  However, we have shown that Ensembl models in chicken are missing a
substantial number of splice variants found in ESTs and mRNAs and Cufflinks does
not recover all splice variants.  A previous study by Schulz \emph{et
al.}\cite{Schulz:2012je} demonstrated that Cufflinks outperformed \emph{de novo}
assembly in detecting genes and isoforms in Ensembl annotations in mouse.
However, we found that Cufflinks almost exclusively finds splice junctions in
mouse Ensembl gene models whereas our gene models include splice junctions not in
Ensembl gene models but supported by ESTs (Figure~\ref{mus_venn}).

To build mouse Cufflinks gene models, we used Tophat default parameters claimed
to be fine-tuned for mammalian RNA-Seq reads, but Cufflinks still poorly
detected splice junctions not in Ensembl.  This suggests that the parameters may
have been set based on mammalian Ensembl models, which could limit the
efficiency of finding novel splice variants in other non-mammal organisms.  In
contrast, {\em de novo} assembly does not rely on species-specific parameters.
It is, therefore, recommended that a combination of Cufflinks models and
transcripts from {\em de novo assembly} should be used to build gene models that
include more splice variants.

We have also shown that the local assembly could greatly increase sensitivity
of splice variant detection. However, how the local assembly affects the
assembly process is not clearly understood.  We speculate that reads with
multiple alignments may play a significant role in enhancing the assembly of
splice variants.  Understanding of the mechanisms may lead to a technique that
can be applied to find splice variants in organisms that lack a reference
genome.

Both Cufflinks and our pipeline rely heavily on a reference genome; therefore,
the quality of the genome will greatly affect the quality of the gene models.
In this study gene models were built from chicken genome version 2.1 (galGal3),
which contains ~17 Mb of sequence duplications and missassemblies
that were eliminated in the latest version of genome assembly
(galGal4).  Duplications and misassemblies lead to false splice junctions, which
in turn produce a large number of splice variants as observed in some
chromosomes.

@@

\subsection*{Mapping reads to The Genome and Gene Models}

Single and paired-end reads were mapped to the chicken genome by
Tophat2\cite{Trapnell:2009dp} release 2.0.0 using default parameters without
annotations.  All reads were mapped to cDNA sequences derived from gene models
by Bowtie2\cite{Langmead:2009fv} with default parameters \texttt{(n=2, l=28,
e=70, k=1)}.  Reads from the mouse dataset were mapped to the mouse genome
(mm9) downloaded from Tophat website \texttt{http://tophat.cbcb.umd.edu}.

@@

Reads from each dataset were first assembled separately by global assembly
without using a reference genome.  In contrast, reads from each dataset were
first mapped to the chicken genome using Tophat2.  Then only reads mapped to
the genome were assembled by chromosome in the local assembly
(Fig.~\ref{local_assembly}).  Global and local assembly was performed using
Velvet version 1.2.03\cite{Zerbino:2008vu} with default parameters except for
hash length (k-mer).  A range of k-mer length from 21-31 was used to assemble
reads from chicken data and k-mer length 27 was used to assemble reads from
mouse data.  Lastly, transcripts from both methods were assembled by Oases
version 0.2.06\cite{Schulz:2012je}.  A poly-A tail, short transcripts and
transcripts with low complexity are removed by seqclean~\cite{seqclean} with
default parameters.  Redundant transcripts are removed by cd-hit-est from the
CD-HIT suite\cite{Li:2006hr}.  A substantial number of transcripts are removed
at this step, which facilitates gene model construction process.

@@

Figure~\ref{overall_pipeline} depicts an overall gene model construction
pipeline.  Transcripts of all datasets from local and global assembly were
mapped to the chicken genome using BLAT\cite{Kent:2002tv} \texttt{(-t=dna -q=dna
-noHead -out=psl -mask=lower -extendThroughN -dots=1000)}.  Alignments and gaps
from BLAT outputs are considered exons and introns respectively.  Optionally,
data from other sources (ESTs, RefGenes, etc.) can be incorporated with
transcripts from the assembly to improve gene models.  All transcripts are then
assembled using Gimme, a program that assembles transcripts based on their
alignments to the reference genome.  An algorithm for assembling transcripts is
described below.  A maximum set of transcripts obtained from Gimme are then
reduced to only a minimum set of transcripts that contain all splice junctions
and untranslated regions (UTRs).  After that, transcripts that are highly
similar ($>99\%$) are clustered and removed by CD-HIT version
4.5.6\cite{Li:2006hr}.  Parameters for clustering are \texttt{-n 10, -r 0 and -c
0.99}.  Only a representative of each cluster is kept in gene models.

@@

A gene model can be represented as a splice graph composed of exons as nodes and
introns as edges.  However, transcripts of the same gene vary in size and
structure depending on the expression level and a hash length number used in
the assembly.  Furthermore, incomplete exons and fragmented transcripts complicate
the construction of a splice graph.  In this study, we developed an algorithm
that handles incomplete exons and fragmented transcripts and constructs a
maximum assembly of gene models.

The algorithm first builds an intron graph using introns as nodes.  Each intron
contains exons one of whose splice sites perfectly match intron boundaries.
Exons are considered incomplete and eliminated if they locate at the $3'$ or
$5'$ end of the transcripts and they are not the largest exons.  ,
Fig.~\ref{algorithm}).  Transcripts were then grouped into the same gene if
they have at least one intron or exon in common.  Then, a splice graph composed
of exons is created and structures of isoforms are derived from traversing
paths in the splice graph.  Gimme is open-source and available at
\texttt{https://github.com/ged-lab/gimme}.

@@

We employed ESTScan version 3.0.3 to translate protein sequences from our gene
models.  The matrix used for building Hidden Markov models was built from chicken
reference cDNA sequences using tools from ESTScan. Only protein sequences longer
than 50 bp are included in the analysis.

@@

\subsection*{Sequence assembly using Cufflinks}
Reads are mapped to a genome sequence using Tophat.  Gene models are built from
each dataset by Cufflinks 2.0.0\cite{Trapnell:2010kd}.  All gene models are
then merged together using Cuffmerge.

\subsection*{Expressed sequence tags and Genbank mRNA}
Expressed sequence tags (ESTs) and mRNAs were downloaded from the UCSC genome
website.  The database was loaded from GENBANK on 1 January 2014.  Sequences
were aligned to the chicken genome using BLAT.

@@

\section*{Figure Legends}
\begin{figure}[!ht]
\begin{center}
\includegraphics[width=5in]{overall_pipeline.pdf}
\end{center}
\caption{
{\bf Gene model construction pipeline.} Transcripts are obtained from two
assembly methods -- global and local assembly.  Transcripts are aligned to the
chicken genome by BLAT\@. Gimme then constructs gene models based on alignments
of transcripts.  }
\label{overall_pipeline}
\end{figure}
